*Supplement* Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2024-01-22 manu-draft.qmd
replace rainfall tia data with swfwmd data, reevaluate trend to rainy season #2

Two primary sources of data have been used to track seagrass change over time in Tampa Bay. The Southwest Florida Water Management District has estimated areal coverage of seagrasses approximately biennially since the late 1980s (Figure [-@fig-seagrasschg]a, available at <https://data-swfwmd.opendata.arcgis.com/>). These maps are created by photointerpretation of aerial images obtained at the end of the growing season, typically during November-February. No species information is provided. A more detailed, but spatially-specific, data source is the Tampa Bay Interagency Seagrass Monitoring Program (Figure [-@fig-seagrasschg]b, <https://tampabay.wateratlas.usf.edu/seagrass-monitoring/>). Annual transect surveys have been conducted since 1998 at 62 fixed locations in Tampa Bay, many of which were chosen to target seagrass beds of interest [@tbep0816; @sherwood2017]. This dataset provides species information on abundance, cover, frequency occurrence, and condition, collected at fixed meter marks along a transect extending from the shoreline to the deepwater edge of the seagrass bed. Although the areal maps provide the standard for assessment of restoration goals, the transect data allow inter-annual comparison at greater temporal resolution, particularly for the recent period of interest when seagrasses have declined. As such, the transect data were used below for comparison with temperature and salinity changes for the major bay segments. Additional sources of seagrass data are described in the next section.

@@

The first goal of the analysis was to describe spatial and temporal trends in water temperature and salinity using the three water quality datasets described above. This assessment provided an indication of the extent of change in Tampa Bay as context for understanding potential relationships with seagrass change. An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions related to long-term (multi-decadal) climate change. For comparison to water quality conditions, meteorological data describing air temperature and precipitation were obtained from Tampa International Airport where daily measurements have been collected since 1939. The *rnoaa* R package [@chamberlain23] was used to retrieve these data starting from 1975 when EPC monitoring began in Tampa Bay. Annual averages for air temperature and cumulative annual precipitation were calculated. Additionally, the Standardized Precipitation Index [SPI, @begueria2013] was estimated from the daily rainfall data to identify periods of time when rainfall significantly deviated from the long-term average [using the *spei* R package, @begueria23]. Annual hydrologic loading data to Tampa Bay beginning in 1985 were also obtained for comparison to annual precipitation [@tbep0623]. All climate and loading data were evaluated annually with simple linear regression trends to assess change over time. Water temperature and salinity trends using the EPC, FIM, and PDEM data were similarly evaluated by averaging the monthly data each year for each bay segment. Linear trends for these data were evaluated based on averages of all stations and months within each year and bay segment, which allowed for comparable statistical power between datasets with different sample designs.

@@

prp <- speidat %>% 
  select(yr, precip_mm) %>% 
  summarise(prptot = sum(precip_mm), .by = yr)
prpmod <- lm(prptot ~ yr, prp)
prpslo <- round(coefficients(summary(prpmod))[2, 1], 2)
prppvl <- p_txt(coefficients(summary(prpmod))[2, 4])
prprsq <- round(summary(prpmod)$adj.r.squared, 2)
prpprd <- predict(prpmod, se = T)
prpstr <- paste0(round(prpprd$fit[1], 1), ' (+/-', round(prpprd$se.fit[1], 2), ' st. err.)')
prpend <- paste0(round(prpprd$fit[nrow(prp)], 1), ' (+/-', round(prpprd$se.fit[nrow(prp)], 2), ' st. err.)')
prpdif <- round(prpprd$fit[nrow(prp)]- prpprd$fit[1], 1)
```

Long-term meteorological data from Tampa International Airport showed significantly increasing trends for air temperature and precipitation ([@fig-meteowqraw]). Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtemppvl`, $R^2$ `r airtemprsq`). Mean annual air temperature in 1975 was `r airtempstr` $^\circ$C, whereas current mean annual air temperature in 2022 was `r airtempend` $^\circ$C, showing an overall increase in the period of record of `r airtempdif` $^\circ$C. Similarly, total precipitation has increased by `r prpslo` mm per year (`r prppvl`, $R^2$ `r prprsq`). Total precipitation in 1975 was `r prpstr` mm, whereas current total precipitation in 2022 was `r prpend` mm, showing an overall increase in the period of record of `r prpdif` mm. Trends in total annual hydrologic load from 1985 to present followed total annual precipitation, although the linear model was not significant likely due to a shorter period of record. The SPI showed notable anomalies in precipitation, with pronounced rainy periods in the early 1980s, late 1990s, 2005, and the last ten years.

@@

![Long-term air temperature, precipitation, hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1975 to 2022. Points for water temperature and salinity are colored by sampling location in the water column and show the average (95% confidence interval) across all stations in each bay segment and sampling months for each year. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/meteowqraw.png){#fig-meteowqraw}

>>>>

Two primary sources of data have been used to track seagrass change over time in Tampa Bay. The Southwest Florida Water Management District (SWFWMD) has estimated areal coverage of seagrasses approximately biennially since the late 1980s (Figure [-@fig-seagrasschg]a, available at <https://data-swfwmd.opendata.arcgis.com/>). These maps are created by photointerpretation of aerial images obtained at the end of the growing season, typically during November-February. No species information is provided. A more detailed, but spatially-specific, data source is the Tampa Bay Interagency Seagrass Monitoring Program (Figure [-@fig-seagrasschg]b, <https://tampabay.wateratlas.usf.edu/seagrass-monitoring/>). Annual transect surveys have been conducted since 1998 at 62 fixed locations in Tampa Bay, many of which were chosen to target seagrass beds of interest [@tbep0816; @sherwood2017]. This dataset provides species information on abundance, cover, frequency occurrence, and condition, collected at fixed meter marks along a transect extending from the shoreline to the deepwater edge of the seagrass bed. Although the areal maps provide the standard for assessment of restoration goals, the transect data allow inter-annual comparison at greater temporal resolution, particularly for the recent period of interest when seagrasses have declined. As such, the transect data were used below for comparison with temperature and salinity changes for the major bay segments. Additional sources of seagrass data are described in the next section.

@@

The first goal of the analysis was to describe spatial and temporal trends in water temperature and salinity using the three water quality datasets described above. This assessment provided an indication of the extent of change in Tampa Bay as context for understanding potential relationships with seagrass change. An assumption was that any changes in physical characteristics in Tampa Bay were driven by interannual changes in weather conditions related to long-term (multi-decadal) climate change. For comparison to water quality conditions, daily air temperature data were obtained from Tampa International Airport [using the *rnoaa* R package, @chamberlain23] and monthly rainfall data for the Tampa Bay watershed were obtained from the SWFWMD (<https://www.swfwmd.state.fl.us/resources/data-maps/rainfall-summary-data-region>). Only rainfall data for the rainy season (June to September) were evaluated. Additionally, the Standardized Precipitation Index [SPI, @begueria2013] was estimated from the rainfall data to identify periods of time when rainfall significantly deviated from the long-term average [using the *spei* R package, @begueria23]. Annual hydrologic loading data to Tampa Bay beginning in 1985 were also obtained for comparison to annual precipitation [@tbep0623]. All climate and loading data were evaluated annually with simple linear regression trends to assess change over time. Water temperature and salinity trends using the EPC, FIM, and PDEM data were similarly evaluated by averaging the monthly data each year for each bay segment. Linear trends for these data were evaluated based on averages of all stations and months within each year and bay segment, which allowed for comparable statistical power between datasets with different sample designs.

@@

prp <- speidat %>% 
  select(yr, mo, precip_mm) %>%
  filter(mo %in% 6:9) %>% 
  summarise(prptot = sum(precip_mm), .by = yr)
prpmod <- lm(prptot ~ yr, prp)
prpslo <- round(coefficients(summary(prpmod))[2, 1], 2)
prppvl <- p_txt(coefficients(summary(prpmod))[2, 4])
prprsq <- round(summary(prpmod)$adj.r.squared, 2)
prpprd <- predict(prpmod, se = T)
prpstr <- paste0(round(prpprd$fit[1], 1), ' (+/-', round(prpprd$se.fit[1], 2), ' st. err.)')
prpend <- paste0(round(prpprd$fit[nrow(prp)], 1), ' (+/-', round(prpprd$se.fit[nrow(prp)], 2), ' st. err.)')
prpdif <- round(prpprd$fit[nrow(prp)]- prpprd$fit[1], 1)

prp2 <- speidat %>% 
  select(yr, mo, precip_mm) %>%
  filter(mo %in% 6:8) %>% 
  summarise(prptot = sum(precip_mm), .by = yr)
prpmod2 <- lm(prptot ~ yr, prp2)
prpslo2 <- round(coefficients(summary(prpmod2))[2, 1], 2)
prppvl2 <- p_txt(coefficients(summary(prpmod2))[2, 4])
prprsq2 <- round(summary(prpmod2)$adj.r.squared, 2)
prpprd2 <- predict(prpmod2, se = T)
prpstr2 <- paste0(round(prpprd2$fit[1], 1), ' (+/-', round(prpprd2$se.fit[1], 2), ' st. err.)')
prpend2 <- paste0(round(prpprd2$fit[nrow(prp)], 1), ' (+/-', round(prpprd2$se.fit[nrow(prp)], 2), ' st. err.)')
prpdif2 <- round(prpprd2$fit[nrow(prp)] - prpprd2$fit[1], 1)
```

Long-term meteorological data showed increasing trends for air temperature and precipitation ([@fig-meteowqraw]). Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtemppvl`, $R^2$ `r airtemprsq`). Mean annual air temperature in 1975 was `r airtempstr` $^\circ$C, whereas current mean annual air temperature in 2022 was `r airtempend` $^\circ$C, showing an overall increase in the period of record of `r airtempdif` $^\circ$C. Similarly, total precipitation during the rainy season has increased by `r prpslo` mm per year, although the model was not significant (`r prppvl`, $R^2$ `r prprsq`). Removing September from the rainy season showed a significantly increasing trend of `r prpslo2` mm per year (`r prppvl2`, $R^2$ `r prprsq2`). Using this model, total precipitation in 1975 was `r prpstr2` mm, whereas current total precipitation in 2022 was `r prpend2` mm, showing an overall increase in the period of record of `r prpdif2` mm. Trends in total annual hydrologic load from 1985 to present followed total annual precipitation, although the linear model was not significant likely due to a shorter period of record. The SPI showed notable anomalies in precipitation, with pronounced rainy periods in the early 1980s, late 1990s, 2005, and 2015-2020.

@@

![Long-term air temperature, precipitation (Jun-Aug), hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1975 to 2022. Points for water temperature and salinity are colored by depth and shown as the average (95% confidence interval) across all stations in each bay segment and sampling months for each year. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/meteowqraw.png){#fig-meteowqraw}

### 2024-01-22 supplement.qmd
replace rainfall tia data with swfwmd data, reevaluate trend to rainy season #2

![Air temperature, precipitation, hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1998 to 2022. Points for water temperature and salinity are colored by sampling location in the water column and show the average (95% confidence interval) across all stations in each bay segment and sampling months for each year. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppmeteowqraw.png){#fig-suppmeteowqraw}

>>>>

![Air temperature, precipitation (Jun-Aug), hydrologic load, Standard Precipitation Index (SPI), water temperature, and salinity trends from 1998 to 2022. Points for water temperature and salinity are colored by depth and shown as the average (95% confidence interval) across all stations in each bay segment and sampling months for each year. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/suppmeteowqraw.png){#fig-suppmeteowqraw}
